\chapter*{Acknowledgements}
\markboth{Acknowledgements}{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

First and foremost, I would like to thank my thesis director Giuseppe Carleo, who has always been a pioneering figure in the field of neural quantum states and guided me through my four years of PhD study.
This journey would have been impossible to complete without the generous help of my postdoc colleagues --- Riccardo Rossi, who can always ask questions; Filippo Vicentini, who knows NetKet and typography; Indaco Biazzo, who knows replica symmetry breaking; Fan Yang, who knows solid-state theory; and the valuable experiences from Jannes Nys, Friederike Metz, and Zakari Denis.
I am equally thankful to the joyful discussions with my PhD colleagues and office mates --- Stefano Barison, Julien Gacon, Gabriel Pescia, Imelda Romero, Alessandro Sinibaldi, Gian Gentinetta, Linda Mauron, David Linteau, Douglas Hendry, Sara Santos, and particularly --- Clemens Giuliani, who is a bit less quiet than me and has helped me a lot with the hardware problems of my workstation in the office.
Moreover, I am thankful to Svetlana Mashkina, who has kindly handled all the secretary works for us.

I would also like to thank the collaborators I have met, and more researchers from whom I have learned.
Attila Szabó, Christopher Roth, and Damian Hofmann have given me great help in the development of NetKet.
Juan Carrasquilla, Mohamed Hibat-Allah, and Or Sharir have shown me different routes in the search space of neural quantum states.
Federico Becca has led me a thoughtful discussion on resonating valence bonds states, and Luciano Viteritti has provided me the latest insights on vision transformers.
Tom Westerhout's lattice-symmetries is the fastest exact diagonalization library I have seen. It is written in Haskell, and it is hard for one to write a program with ever a bug in Haskell.
I am also thankful to the friends I have made in EPFL --- Jin Jiang, Yifei Guan, and Longdi Zhu, who have helped me solve the problem that the dishes in Chinese restaurants are too much for one person.

Finally, I would like to thank my parents, with whom I spend most time on video calling rather than text messaging.
I am thankful as well to the indie game dev group Lunatic Works and the indie AI dev group ChatGal, who have fulfilled my spare time and have proven that friendship is not hindered by geographical distance.

\begin{flushright}
Dian Wu \\
November 13, 2024
\end{flushright}
